Host richness increases the occurrence but not the severity of bark beetle-induced tree mortality

### 2021-11-22 Code/Manuscript-many.Rmd
including multiple ways for determining presence

df$hostsPICO25 <- ifelse(df$BA.PICO/df$BA.total>0.25, "PICO", "absent")
df$hostsPIEN25 <- ifelse(df$BA.PIEN/df$BA.total>0.25, "PIEN", "absent")
df$hostsABLA25 <- ifelse(df$BA.ABLA/df$BA.total>0.25, "ABLA", "absent")
df$hosts25 <-  do.call(paste, c(df[c("hostsPICO25", "hostsPIEN25", "hostsABLA25")],sep="-"))
df$hosts25 <- factor(df$hosts25, levels=c('PICO-absent-absent','absent-absent-ABLA', 'absent-PIEN-absent', 'PICO-PIEN-absent',   'PICO-absent-ABLA', 'absent-PIEN-ABLA', 'PICO-PIEN-ABLA', 'absent-absent-absent'), labels= c('PICO', 'ABLA', 'PIEN', 'PICO+PIEN', 'PICO+ABLA', 'PIEN+ABLA', 'PICO+PIEN+ABLA', 'none'), ordered=TRUE)

df$nhosts25 <- df$hosts25
df$nhosts25 <- factor(df$nhosts25, levels=c('PICO', 'ABLA', 'PIEN', 'PICO+PIEN', 'PICO+ABLA', 'PIEN+ABLA', 'PICO+PIEN+ABLA', 'none'), labels=c(1, 1,1,2,2,2,3,0))

write.csv(df, here("Data/Processed/FIAdat-df.csv"),row.names=F)
```

@@

write.csv(df, here("Data/Processed/FIAdat-recent-re.csv"),row.names=F)
```

```{r import processed data}
df <- read.csv( here("Data/Processed/FIAdat-recent-re.csv"))
df$nhostsF <-factor(df$nhosts, levels=1:3, labels=c("one", "two", "three"), ordered=T)
df$hosts <- factor(df$hosts, levels=c("PICO", "PIEN", "ABLA", "PICO+PIEN", "PICO+ABLA", "PIEN+ABLA", "PICO+PIEN+ABLA"), ordered=T)
df$hostsF <- factor(df$hosts, levels=c("PICO", "PIEN", "ABLA", "PICO+PIEN", "PICO+ABLA", "PIEN+ABLA", "PICO+PIEN+ABLA"), labels=c("PICO", "PIEN", "ABLA", "PICO & PIEN", "PICO & ABLA", "PIEN & ABLA", "PICO, PIEN, & ABLA"), ordered=T)

df$nhosts25F <-factor(df$nhosts, levels=1:3, labels=c("one", "two", "three"), ordered=T)
df$hosts25 <- factor(df$hosts, levels=c("PICO", "PIEN", "ABLA", "PICO+PIEN", "PICO+ABLA", "PIEN+ABLA", "PICO+PIEN+ABLA"), ordered=T)
df$hosts25F <- factor(df$hosts, levels=c("PICO", "PIEN", "ABLA", "PICO+PIEN", "PICO+ABLA", "PIEN+ABLA", "PICO+PIEN+ABLA"), labels=c("PICO", "PIEN", "ABLA", "PICO & PIEN", "PICO & ABLA", "PIEN & ABLA", "PICO, PIEN, & ABLA"), ordered=T)

@@

```{r maps all plots and hosts, eval=F}
df.all <- df[df$focalhost == "Any", ]
df.all$agentsFlong <-factor(df.all$agentsF, levels=levels(df.all$agentsF), labels=c("unaffected", "MPB", "SB", "WBBB", "MPB & SB", "MPB & WBBB", "SB & WBBB", "MPB, SB, & WBBB"))

@@

map1 <- tm_shape(IMW, bbox=tmaptools::bb(ext)) +  tm_fill(col=grey(level=0.95)) +  tm_borders()  + tm_shape(df.all.sf[df.all.sf$hosts=="PICO",]) +tm_dots(col="agentsFlong", legend.show = FALSE,size=.0025, palette=c(unaffected="#A6A6A6", MPB="#710002")) +tm_compass(position=c(0.1, 0.02), just=c("center", "bottom"), text.size=0.5, size=2)+tm_layout(main.title = "PICO",main.title.size=0.5,main.title.position="center")

map2 <- tm_shape(IMW, bbox=tmaptools::bb(ext)) +  tm_fill(col=grey(level=0.95)) +  tm_borders()  + tm_shape(df.all.sf[df.all.sf$hosts=="PIEN",]) +tm_dots(col="agentsFlong", legend.show = FALSE,size=.0025, palette=rev(p2col)) +tm_layout(main.title = "PIEN",main.title.size=0.5,main.title.position="center")

map3 <- tm_shape(IMW, bbox=tmaptools::bb(ext)) +  tm_fill(col=grey(level=0.95)) +  tm_borders()  + tm_shape(df.all.sf[df.all.sf$hosts=="ABLA",]) +tm_dots(col="agentsFlong", legend.show = FALSE,size=.0025, palette=rev(p2col)) +tm_layout(main.title = "ABLA",main.title.size=0.5, main.title.position="center")

map4 <- tm_shape(IMW, bbox=tmaptools::bb(ext)) +  tm_fill(col=grey(level=0.95)) +  tm_borders()  + tm_shape(df.all.sf[df.all.sf$hosts=="PICO+PIEN",]) +tm_dots(col="agentsFlong", legend.show = FALSE,size=.0025, palette=rev(p2col)) +tm_layout(main.title = "PICO & PIEN",main.title.size=0.5,main.title.position="center")

map5 <- tm_shape(IMW, bbox=tmaptools::bb(ext)) +  tm_fill(col=grey(level=0.95)) +  tm_borders()  + tm_shape(df.all.sf[df.all.sf$hosts=="PICO+ABLA",]) +tm_dots(col="agentsFlong", legend.show = FALSE,size=.0025, palette=rev(p2col)) +tm_layout(main.title = "PICO & ABLA", main.title.size=0.5, main.title.position="center")

map6 <- tm_shape(IMW, bbox=tmaptools::bb(ext)) +  tm_fill(col=grey(level=0.95)) +  tm_borders()  + tm_shape(df.all.sf[df.all.sf$hosts=="PIEN+ABLA",]) +tm_dots(col="agentsFlong", legend.show = FALSE,size=.0025, palette=rev(p2col)) +tm_layout(main.title = "PIEN & ABLA", main.title.size=0.5, main.title.position="center")

map7 <- tm_shape(IMW, bbox=tmaptools::bb(ext)) +  tm_fill(col=grey(level=0.95)) +  tm_borders()  + tm_shape(df.all.sf[df.all.sf$hosts=="PICO+PIEN+ABLA",]) +tm_dots(col="agentsFlong", legend.show = FALSE,size=.0025, palette=rev(p2col))+tm_layout(main.title = "PICO, PIEN, & ABLA", main.title.size=0.5, main.title.position="center")

legend.map <- tm_shape(IMW, bbox=tmaptools::bb(ext)) +  tm_fill(col=grey(level=0.95)) +  tm_borders()  + tm_shape(df.all.sf[df.all.sf$hosts=="PICO+PIEN+ABLA",]) +tm_dots(col="agentsFlong", size=4, title="bark beetle species",  palette=rev(p2col))+tm_layout(legend.only = TRUE,legend.text.size = 0.5,legend.title.size = 0.65)+tm_scale_bar(position = c(0.5,0.16), text.size=0.5, just="center", breaks=c(0,250,500))

@@

```{r fit MPB RF model, eval=F,fig.show='hide'}
response <- "agentMPB"
vars <- c("BA.PICO", "QMD.PICO", "propBA.PICO") # predictor variables
datBB.c <- df[df$focalhost=="Any",] # dataframe containing response and predictors (here i had to subset it because of how my data was arranged)

@@

# partial dependence
pdale <- NULL
for(z in vars.final){
  pdz <- partial(ranger_obj, train=datz.train, pred.var=z, which.class=strsplit(response, split="agent")[[1]][2], prob=T)
  pdz$var <- z
  colnames(pdz) <- c("value","yhat","var")
  pdz$method <- "PD"

  alez <- ALEPlot(X=datz.train[,vars.final], X.model=ranger_obj, pred.fun=function(X.model, newdata){predict(X.model, newdata)$predictions[,2]}, J=which(vars.final==z), K=50)
  alez <- data.frame(value=alez$x.values, yhat=alez$f.values, var=z, method="ALE")

  pdale <- rbind(pdale, pdz)
  pdale <- rbind(pdale, alez)
}
write.csv(pdale,  here(paste0("Results/RF-partialdependence-", response, "-occurrence-SMOTE.csv")), row.names=F)
```

>>>>

write.csv(df, here("Data/Processed/FIAdat-df-outbreak.csv"),row.names=F)
```

@@

write.csv(df, here("Data/Processed/FIAdat-df-merge-re.csv"),row.names=F)
```

```{r reorganize by focal host - 25%, eval=F}
df <- read.csv(here("Data/Processed/FIAdat-recent.csv"))

formatp <- function(d, host="PIEN", agent="SB"){
  df.p <- d[which(d[, paste0("TPH.", host)]>0),]
  df.p <- df.p[df.p$BA.host/df.p$BA.total>0.25, ]
  df.p$TPH.host <- df.p[, paste0("TPH.", host)]
  df.p$QMD.host <- df.p[, paste0("QMD.", host)]
  df.p$BA.host <- df.p[, paste0("BA.", host)]

write.csv(df, here("Data/Processed/FIAdat-df-merge-re.csv"),row.names=F)
```

```{r import processed data}
df <- read.csv( here("Data/Processed/FIAdat-df-merge-re.csv"))
df$nhostsF <-factor(df$nhosts, levels=1:3, labels=c("one", "two", "three"), ordered=T)
df$hosts <- factor(df$hosts, levels=c("PICO", "PIEN", "ABLA", "PICO+PIEN", "PICO+ABLA", "PIEN+ABLA", "PICO+PIEN+ABLA"), ordered=T)
df$hostsFP <- factor(df$hosts, levels=c("PICO", "PIEN", "ABLA", "PICO+PIEN", "PICO+ABLA", "PIEN+ABLA", "PICO+PIEN+ABLA"), labels=c("PICO", "PIEN", "ABLA", "PICO & PIEN", "PICO & ABLA", "PIEN & ABLA", "PICO, PIEN, & ABLA"), ordered=T)

@@

```{r maps all plots and hosts, eval=F}
df.all <- df[df$focalhost == "Any", ]
df.all$hostsFPlong <- df.all$hostsFP
df.all$agentsFplong <-factor(df.all$agentsF, levels=levels(df.all$agentsF), labels=c("unaffected", "MPB", "SB", "WBBB", "MPB & SB", "MPB & WBBB", "SB & WBBB", "MPB, SB, & WBBB"))

@@

map1 <- tm_shape(IMW, bbox=tmaptools::bb(ext)) +  tm_fill(col=grey(level=0.95)) +  tm_borders()  + tm_shape(df.all.sf[df.all.sf$hostsFPlong=="PICO",]) +tm_dots(col="agentsFplong", legend.show = FALSE,size=.0025, palette=c(unaffected="#A6A6A6", MPB="#710002")) +tm_compass(position=c(0.1, 0.02), just=c("center", "bottom"), text.size=0.5, size=2)+tm_layout(main.title = "PICO",main.title.size=0.5,main.title.position="center")

map2 <- tm_shape(IMW, bbox=tmaptools::bb(ext)) +  tm_fill(col=grey(level=0.95)) +  tm_borders()  + tm_shape(df.all.sf[df.all.sf$hostsFPlong=="PIEN",]) +tm_dots(col="agentsFplong", legend.show = FALSE,size=.0025, palette=rev(p2col)) +tm_layout(main.title = "PIEN",main.title.size=0.5,main.title.position="center")

map3 <- tm_shape(IMW, bbox=tmaptools::bb(ext)) +  tm_fill(col=grey(level=0.95)) +  tm_borders()  + tm_shape(df.all.sf[df.all.sf$hostsFPlong=="ABLA",]) +tm_dots(col="agentsFplong", legend.show = FALSE,size=.0025, palette=rev(p2col)) +tm_layout(main.title = "ABLA",main.title.size=0.5, main.title.position="center")

map4 <- tm_shape(IMW, bbox=tmaptools::bb(ext)) +  tm_fill(col=grey(level=0.95)) +  tm_borders()  + tm_shape(df.all.sf[df.all.sf$hostsFPlong=="PICO & PIEN",]) +tm_dots(col="agentsFplong", legend.show = FALSE,size=.0025, palette=rev(p2col)) +tm_layout(main.title = "PICO & PIEN",main.title.size=0.5,main.title.position="center")

map5 <- tm_shape(IMW, bbox=tmaptools::bb(ext)) +  tm_fill(col=grey(level=0.95)) +  tm_borders()  + tm_shape(df.all.sf[df.all.sf$hostsFPlong=="PICO & ABLA",]) +tm_dots(col="agentsFplong", legend.show = FALSE,size=.0025, palette=rev(p2col)) +tm_layout(main.title = "PICO & ABLA", main.title.size=0.5, main.title.position="center")

map6 <- tm_shape(IMW, bbox=tmaptools::bb(ext)) +  tm_fill(col=grey(level=0.95)) +  tm_borders()  + tm_shape(df.all.sf[df.all.sf$hostsFPlong=="PIEN & ABLA",]) +tm_dots(col="agentsFplong", legend.show = FALSE,size=.0025, palette=rev(p2col)) +tm_layout(main.title = "PIEN & ABLA", main.title.size=0.5, main.title.position="center")

map7 <- tm_shape(IMW, bbox=tmaptools::bb(ext)) +  tm_fill(col=grey(level=0.95)) +  tm_borders()  + tm_shape(df.all.sf[df.all.sf$hostsFPlong=="PICO, PIEN, & ABLA",]) +tm_dots(col="agentsFplong", legend.show = FALSE,size=.0025, palette=rev(p2col))+tm_layout(main.title = "PICO, PIEN, & ABLA", main.title.size=0.5, main.title.position="center")

@@

```{r fit MPB RF model, eval=F}
response <- "agentMPB"
vars <- c("BA.PICO", "QMD.PICO", "propBA.PICO") # predictor variables
datBB.c <- df[df$focalhost=="Any",] # dataframe containing response and predictors (here i had to subset it because of how my data was arranged)

@@

# partial dependence
pd <- NULL
for(z in vars.final){
  pdz <- partial(ranger_obj, train=datz.train, pred.var=z, which.class=strsplit(response, split="agent")[[1]][2], prob=T)
  pdz$var <- z
  colnames(pdz) <- c("value","yhat","var")
  pd <- rbind(pd, pdz)
}
write.csv(pd,  here(paste0("Results/RF-partialdependence-", response, "-occurrence-SMOTE.csv")), row.names=F)
```
